---
title: "Data Structures and"
author: "JS"
format: html
editor: visual
---

# Data structures and Data Wrangling

Apart from variables and vectors, there are 4 more data structures

-   factors

-   matrix

-   data frames

-   lists

## Factors

This is a specialized version of a vector oftentimes used to store categorical data. It consists of the underlying vector of values and a separate argument called levels, which represents the different values in the factor.

```{r}
f  <- factor(c("C", "C", "C", "A", "A", "A"),
             levels = c("A", "C"))

sort(f)
f <- relevel(f, "C")

sort(f)
```

## Matrix

```{r}
m <- matrix(1:20, nrow = 5)
m
m[1,]
m[,1]
colnames(m) <- c("a", "b", "c", "d")
m
rownames(m) <- c("v", "w", "x", "y", "z")
m

m["x", "a"]
```

## Data Frame

```{r}
set.seed(259)
df <- data.frame(diagnosis = rep(c("CTRL", "AAA"), each = 5), 
                 diameter = c(rnorm(5, 5, 0.5), 
                              rnorm(5, 2, 0.1)),
                 age = round(runif(10, 45, 80)))
df

df[,1]
df$diameter
summary(df)


df$diagnosis <- as.factor(df$diagnosis)

summary(df)
```

## List

```{r}
examples <- list("a", c("c", "example"), df)
examples[[1]]


examples[1:2]


length(examples)
str(examples)

```

### Named lists

```{r}
ducks <- list(number_of_ducks = 3,
              duck_names = c("Tick", "Trick", "Track"),
              duck_info = df)
ducks[[1]]
ducks[["duck_info"]]
ducks$duck_names
```

```{r}

```
