---
title: "solutions 2"
author: "JS"
format: 
  html:
    self-contained: true
    toc: true
    toc-depth: 4
    df-print: paged
    grid: 
      body-width: 1200px
editor: visual
---

```{r}
library(tidyverse)
library(here)
```

# Easy

## E1: Subset the dataset to include only contain 3 observations per diagnosis. Create a box plot of abdominal aortic diameter. Repeat the sub-setting and the plotting. Is the plot changing?

```{r}
df <- read_tsv(here::here("data/aneurysm_survey/example_data.txt"))
```

```{r}
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 3) %>% 
  ungroup() %>% 
  ggplot(aes(diagnosis, abdominal_diameter))+
  geom_boxplot()

df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 3) %>% 
  ungroup() %>% 
  ggplot(aes(diagnosis, abdominal_diameter))+
  geom_boxplot()
```

## E2: Repeat this with 10 observations per diagnosis. How stable is the re-sampling then?

```{r}
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 10) %>% 
  ungroup() %>% 
  ggplot(aes(diagnosis, abdominal_diameter))+
  geom_boxplot()

df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 10) %>% 
  ungroup() %>% 
  ggplot(aes(diagnosis, abdominal_diameter))+
  geom_boxplot()
```

## E3: Lets stick with the 10 observations per observations. Create a boxplot again. Add points on top of it.

```{r}

df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 10) %>% 
  ungroup() %>% 
  ggplot(aes(diagnosis, abdominal_diameter))+
  geom_boxplot()+
  geom_point()
```

## E4: What is the difference between `fill` and `color` for geom_boxplot? Try it out by adding some color to your boxplots.

```{r}
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 10) %>% 
  ungroup() %>% 
  ggplot(aes(diagnosis, abdominal_diameter, color = diagnosis))+
  geom_boxplot()+
  geom_point()
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 10) %>% 
  ungroup() %>% 
  ggplot(aes(diagnosis, abdominal_diameter, fill = diagnosis))+
  geom_boxplot()+
  geom_point()

```

## E5: Create a scatter plot using BMI and abdominal aortic diameter. Do you think there is connection?

```{r}
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 10) %>% 
  ungroup() %>% 
  ggplot(aes(bmi, abdominal_diameter, fill = diagnosis))+
  geom_point()
df %>% 
  # group_by(diagnosis) %>% 
  # slice_sample(n = 10) %>% 
  # ungroup() %>% 
  ggplot(aes(bmi, abdominal_diameter, fill = diagnosis))+
  geom_point()
```

## E6: Reuse the code for the scatter plot. Set the alpha of the points to 0.5 by using `geom_point(alpha = 0.5)`. What happens? How does the plot change when you change the value?

```{r}
df %>% 
  # group_by(diagnosis) %>% 
  # slice_sample(n = 10) %>% 
  # ungroup() %>% 
  ggplot(aes(bmi, abdominal_diameter, fill = diagnosis))+
  geom_point(alpha = 0.5)
df %>% 
  # group_by(diagnosis) %>% 
  # slice_sample(n = 10) %>% 
  # ungroup() %>% 
  ggplot(aes(bmi, abdominal_diameter, fill = diagnosis))+
  geom_point(alpha = 0.2)
```

## E7: Use the same process as above to try some values for the shape of `geom_point()`. The vlaues range from 0 to 25. Which one is a square with a cross in it?

```{r}
df %>% 
  ggplot(aes(bmi, abdominal_diameter))+
  geom_point(shape = 12)
df %>% 
  ggplot(aes(bmi, abdominal_diameter))+
  geom_point(shape = 7)
```

## E8: Using `aes()`, set the size of the points to age. What does the plot look like?

```{r}
df %>% 
  ggplot(aes(bmi, abdominal_diameter, size = age))+
  geom_point(shape = 7)
```

## E9: Try setting the shape of the points by age. What is going wrong?

```{r}
#| error: true

df %>% 
  ggplot(aes(bmi, abdominal_diameter, shape = age))+
  geom_point()
```

## E10: Filter the original data set to patients with a BMI larger than 20. How many patients remain per diagnosis? Plot a bar chart to visualise it.

```{r}
df %>% 
  filter(bmi > 20) %>% 
  ggplot(aes(diagnosis))+
  geom_bar()
```

# Medium

## M1: Create a box plot of abdominal aortic diameter. Reorder the X axis to have the CTRL diagnosis be the reference level

```{r}
df %>% 
  ggplot(aes(diagnosis, abdominal_diameter))+
  geom_boxplot()
df %>% 
  mutate(diagnosis = fct_relevel(diagnosis, "CTRL")) %>% 
  ggplot(aes(diagnosis, abdominal_diameter))+
  geom_boxplot()
```

## M2: Recreate the scatter plot from above. Color the points by diagnosis.

```{r}
df %>% 
  ggplot(aes(bmi, abdominal_diameter, color = diagnosis))+
  geom_point()
```

## M3: Adjust the color code to something of your liking.

```{r}

df %>% 
  ggplot(aes(bmi, abdominal_diameter, color = diagnosis))+
  geom_point()+
  scale_color_manual(values = c("purple", "limegreen", "lightblue"))
```

## M4: Use geom_histogram() to visualise the distribution of abdominal diameter. Facet using the diagnosis.

```{r}
df %>% 
  ggplot(aes(abdominal_diameter))+
  geom_histogram()+
  facet_wrap(~diagnosis)
```

## M5: Sample your data to 10 observations per diagnosis. Use `stat_summary()` to generate a bar chart of thoracic diameter by diagnosis. Add error bars and individual observations.

```{r}
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 10) %>% 
  ungroup() %>% 
  ggplot(aes(diagnosis, abdominal_diameter))+
  stat_summary(geom = "bar")+
  stat_summary(geom = "errorbar")+
  geom_point()
```

## M6: Building on the previous task, improve the looks of the previous plot like you would for an important presentation. Adjust Axis labels, fix the order on the x axis, give the plot a title. Color the bars by diagnosis. Try to make it as pleasing as possible (in your eyes).

```{r}
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 10) %>% 
  ungroup() %>% 
  mutate(diagnosis = fct_relevel(diagnosis, "CTRL")) %>% 
  ggplot(aes(diagnosis, abdominal_diameter, fill = diagnosis))+
  stat_summary(geom = "bar")+
  stat_summary(geom = "errorbar", width = 0.3, alpha = 0.6)+
  geom_point()+
  scale_fill_manual(values = thematic::okabe_ito(n = 3))+
  theme_light()+
  labs(x = "Diagnosis", y = "Abdominal Aortic Diameter",
       title = "Aortic Diameter by Diagnosis",
       caption = "Bars show mean +- SE",
       fill = "Diagnosis:")+
  theme(legend.position = "bottom")
```

## M7: Sample your data to 3 observations per diagnosis. Test the difference in thoracic diameter between the diagnoses. Is the difference significant? Repeat the sampling and test again. Increase the number of observations per diagnosis to 5 and 10. How do the p-values change?

```{r}
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 3) %>% 
  ungroup() %>% 
  rstatix::t_test(thoracic_diameter ~ diagnosis)
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 3) %>% 
  ungroup() %>% 
  rstatix::t_test(thoracic_diameter ~ diagnosis)
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 3) %>% 
  ungroup() %>% 
  rstatix::t_test(thoracic_diameter ~ diagnosis)


df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 5) %>% 
  ungroup() %>% 
  rstatix::t_test(thoracic_diameter ~ diagnosis)
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 5) %>% 
  ungroup() %>% 
  rstatix::t_test(thoracic_diameter ~ diagnosis)


df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 10) %>% 
  ungroup() %>% 
  rstatix::t_test(thoracic_diameter ~ diagnosis)
df %>% 
  group_by(diagnosis) %>% 
  slice_sample(n = 10) %>% 
  ungroup() %>% 
  rstatix::t_test(thoracic_diameter ~ diagnosis)
```

# Hard

## H1: What happens when you run the following code?

```{r}
df <- read_tsv(here("data", "aneurysm_survey", "example_data.txt"))
df_plot <- df %>% 
  group_by(diagnosis, hypertension) %>% 
  slice_sample(n = 5) %>% 
  ungroup() %>% 
  mutate(hypertension = ifelse(hypertension, "present", "normal bp"))

df_plot  %>% 
  ggplot(aes(hypertension, thoracic_diameter, color = diagnosis))+
  geom_boxplot()
```

## H2: Try adding points to the plot using `geom_point()`. What happens? Can you fix their position? Hint: look at the `position` argument of `geom_point()`.

```{r}
df_plot <- df %>% 
  group_by(diagnosis, hypertension) %>% 
  slice_sample(n = 5) %>% 
  ungroup() %>% 
  mutate(hypertension = ifelse(hypertension, "present", "normal bp"))

df_plot  %>% 
  ggplot(aes(hypertension, thoracic_diameter, color = diagnosis))+
  geom_boxplot()+
  geom_point()

df_plot  %>% 
  ggplot(aes(hypertension, thoracic_diameter, color = diagnosis))+
  geom_boxplot()+
  geom_point(position = position_dodge(width  = 0.75))
```

## H3: Using what we learned last session, build an expression plot of *IL6* using the airway data. Best use `geom_boxplot()` and `geom_point()`.

```{r}
counts <- read_tsv(here::here("data/expression_data/counts.txt")) 
annotation <- read_tsv(here::here("data/expression_data/annotation.txt"))

counts_long <- counts %>% pivot_longer( # take the data and pivot longer 
  cols = -gene, # select the columns to pivot longer, here everything but the gene column (similar to select) 
  names_to = "sample", # put the column names into a new column called sample 
  values_to = "count") # put the corresponding values ion a new column called count ) 

anno_filtered <- annotation %>% select(Run, dex, cell)

df_long <- counts_long %>% inner_join(anno_filtered, # join in the filtered annotation 
                                      by = join_by(sample == Run))

df_long %>% 
  filter(gene == "IL6") %>% 
  ggplot(aes(dex, count))+
  geom_boxplot()+
  geom_point()
```

```{r}
stats <- df_long %>% 
  filter(gene == "IL6") %>% 
  rstatix::t_test(count ~ dex) %>% 
  rstatix::add_xy_position()

df_long %>% 
  filter(gene == "IL6") %>% 
  ggplot(aes(dex, count))+
  geom_boxplot()+
  geom_point()+
  ggpubr::stat_pvalue_manual(stats)
```
